---
title: "Hurricane Trends Along the East Coast"
author: "Elise Boos, Andrew Brantley, Kelsey Husted"
output:
  html_document:
    toc: yes
    df_print: paged
  pdf_document:
    keep_tex: yes
    fig_caption: yes
  bookdown::html_document2:
    df_print: paged
    toc: yes
    fig_caption: yes
subtitle: https://github.com/elise-227/BoosBrantleyHusted_ENV872_EDA_FinalProject.git
geometry: margin=2.54cm
fontsize: 12pt
mainfont: Times New Roman
editor_options:
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage

## List of Tables
- Table \@ref(tab:table1): Data Structure 
- Table \@ref(tab:table2): List of tau and p-values from time series analyses

\newpage

## List of Figures
- Exploratory Analysis                                          
   Figure \@ref(fig:ex-an-FL): Florida Mean Daily Discharge During Hurricane Season                  
   Figure \@ref(fig:ex-an-NC): North Carolina Mean Daily Discharge During Hurricane Season                   
   Figure \@ref(fig:ex-NY): New York Mean Daily Discharge During Hurricane Season
   
- Analysis                       
   Figure \@ref(fig:NY-time-series-analysis): Decomposed Components of New York Time Series                 
   Figure \@ref(fig:NY-time-series-analysis-continued): New York Non-Seasonal Daily Discharge During Hurricane Season                    
   Figure \@ref(fig:NC-time-series-analysis): Decomposed Components of North Carolina Time Series             
   Figure \@ref(fig:NC-time-series-analysis-cont): North Carolina Non-Seasonal Daily Discharge During Hurricane Season                
   Figure \@ref(fig:FL-analysis-time-series): Decomposed Components of Florida Time Series                   
   Figure \@ref(fig:FL-analysis-final-plot): Florida Non-Seasonal Daily Discharge During Hurricane Season       
   Figure \@ref(fig:combined-plot): Comparison Between Three Sites: Non-Seasonal Daily Discharge During Hurricane Season

\newpage
\listoffigures

```{r setup, include=FALSE}
#Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(dygraphs)
library(xts)
library(ggfortify)
require(dplyr)
library(Kendall)
library(tseries)
require(sf)
require(leaflet)
require(mapview)
library(plyr)
mapviewOptions(fgb = FALSE)

# Set your ggplot theme

my_theme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom", legend.text=element_text(size=8), 
        legend.title=element_text(size=10), panel.background = element_rect(fill = "gray95"), 
        panel.grid.major = element_line(size = 0.6, linetype = 'solid', colour = "white"),
        title = element_text(size = 12))

theme_set(my_theme)

# Load your datasets

#New York Data
NY_Discharge.processed <- read.csv("../Data/Processed/NY_DischargeData_Processed.csv", stringsAsFactors = TRUE)
#Format Date
NY_Discharge.processed$Date <- as.Date(NY_Discharge.processed$Date, format = "%Y-%m-%d")

#North Carolina Data
NewRiver.Discharge.PROC <- read.csv("../Data/Processed/NC_Discharge_NewRiver_PROC.csv", stringsAsFactors = TRUE)
# setting date column as date
NewRiver.Discharge.PROC$Date <- as.Date(NewRiver.Discharge.PROC$Date, format = '%Y-%m-%d')

#Florida data
florida_hurricane_1990 <- read.csv("../Data/Processed/florida_hurricane_1990_processed.csv", 
                                   stringsAsFactors = T)
#setting date column as date 
florida_hurricane_1990$Date <- as.Date(florida_hurricane_1990$Date, format = '%Y-%m-%d')
```


# Rationale and Research Questions

Hurricanes are a serious natural disaster that affects millions each year along the East Coast of the United States. Furthermore, hurricanes create severe flooding, dangerous storm surges, and high winds that have long-lasting and devastating consequences on many communities (Schmeltz et al. 2013). Understanding how climate change influences the trends of hurricanes will be informative for strategic planning to protect both property and lives in the future (Mudd et al. 2014). For this analysis we chose to look at discharge data in the vicinity of major rivers in states spanning the East Coast of the US. The objective of the analysis is to gain valuable insight on how locality and intensity of these hurricanes may be changing over the past three decades. 

Discharge data in Florida, North Carolina, and New York will be used to gain a better understanding of how hurricane regimes change across the entire span of the coast through time. Furthermore, the comparative study will reveal if hurricanes are shifting poleward. Our central research questions revolve around investigating any increasing trends in discharge during the hurricane season (June-October) and where these increasing trends are more intense. 

# Research Questions

 **Question 1**: Has there been an overall increase in mean daily discharge during the Atlantic hurricane season from 1990 - 2021?

 **Question 2**: Has one portion of the East Coast of the US experienced a disproportionate increase in hurricane activity, and its associated discharge, compared to other portions commonly affected by these hurricanes?

 **Question 3**: Are the general locations of hurricanes shifting towards the poles in the northern and/or southern hemisphere?

\newpage

# Dataset Information

All datasets used in this analysis were downloaded from the United States Geological Survey (USGS) website, specifically from  historical records of discharge data recorded by field dataloggers. The data was obtained from three states varying in latitude along the East Coast (i.e., Florida, North Carolina, and New York). Stream gage locations were selected based on proximity to major river basins and vulnerable counties susceptible to hurricane impacts. Additionally, gage locations were specifically chosen with sufficient data, going back to at least the 1990s. Similar timeline ranges for the datatsets ensures a time series analysis that is comparable among all three sites. Discharge data was the data type used in this analysis with all datasets having at minimum daily recordings but some having recordings as often as every 30 minutes. The basic data structure is outlined below (Table \@ref(tab:table1)).

```{r table1, echo=FALSE}
Name <- c("Agency", "SiteNumber", "MeanDischarge", "Date", "Month")
Description <- c("USGS","USGS gage number", "Mean daily discharge (cfs)", "Date of collection", "Month of collection")
datast <- data.frame(Name, Description)
colnames(datast) <- c("Column Name", "Description")
  
knitr::kable(datast, caption ="Table summarizing the dataset structure")
```
\newpage

# Data Wrangling

Downloaded datasets included date/time and mean discharge (cfs) columns. After the date column was properly classified as a date, unnecessary columns were removed from the data frame. Next, the date column was separated out to filter for years 1990-2021 and to filter for hurricane months (June-October). The dataset was then saved into the processed folder for each of the research locations. Missing data values were then filled using a linear interpolation and then daily averages of discharge were created in a new, clean column of daily means. After all fo these steps were taken the data was considered to be wrangled and was then ready for initial visualizations. 


\newpage

# Exploratory Analysis 

The exploratory analysis of the data involved creating a map of the USGS gage locations along the East Coast of the U.S. to orient individuals towards the locality of the research. Also, the initial visualizations of the discharge data for each research location-during hurricane seasons-was included in the exploratory analysis from ~1990-2021. After the initial graphs displayed  poorly visualized data, a log scale was applied to the y-axis in order to better present any noticeable trend that may be occurring. Comparing differences between the intensity, frequency, and shifts in these variables over time, permitted researchers to formulate predictions on possible changes in each research location as well as any changes over time. However, to fully understand these shifting trends, the following time series analysis was conducted. 

Below is a map of where each USGS gage site is located along the Atlantic coast of the US.

```{r map of locations, warning=FALSE,echo=FALSE}
site_locations <- read.csv("../Data/Raw/Site_locations.csv", stringsAsFactors = T)
site_locations.sf <- site_locations %>%
st_as_sf(coords = c('Long','Lat'),
         crs=4269)
mapview(site_locations.sf[1,], zcol = "Site.Name", col.regions = "deeppink1") +
mapview(site_locations.sf[2,], zcol = "Site.Name", col.regions = "turquoise4") +
mapview(site_locations.sf[3,], zcol = "Site.Name", col.regions = "forestgreen")
```

Figure \@ref(fig:ex-an-FL) is an initial plot of discharge within hurricane months at the Florida location and shows a potential decrease in discharge over time. 

```{r ex-an-FL, warning=FALSE, echo=FALSE, message=FALSE, fig.align='center',fig.cap='Florida Discharge (cfs) 1990-2021'}
#plot discharge before running time series on the log scale
ggplot(florida_hurricane_1990, aes(x = Date, y = log(Daily_Discharge_clean)))+
  geom_point(color = "deeppink1")+
  geom_smooth(method = lm, color = "black")+
  labs(x = "Date", y = "Mean Daily Discharge (cfs)",
       title = "Mean Daily Discharge During Hurricane Seasons (1990-2021)", 
       subtitle = "North Prong St. Sebastian River, Florida") 
```

The following plot (Figure \@ref(fig:ex-an-NC)) for the North Carolina daily discharge dataset displays a potential increasing trend over time.

```{r ex-an-NC, warning=FALSE, echo=FALSE, message=FALSE, fig.align='center', fig.cap='North Carolina Discharge (cfs) 1990-2021'}
# initial data visualiation
DailyDischarge.Plot <- 
                ggplot(NewRiver.Discharge.PROC, aes(Date, MeanDailyDischarge_cfs)) +
                      geom_point(color = "forestgreen") +
                      geom_smooth(method = "lm", color = "black") +
                      labs(x = "Date", y = "Mean Daily Discharge (cfs)",
                        title = "Mean Daily Discharge During Hurricane Seasons (1990-2021)", 
                           subtitle = "New River, North Carolina") +
                      theme(plot.title=element_text(hjust=0.5)) +
                      theme(plot.subtitle=element_text(hjust=0.5)) +
                      scale_y_log10()
DailyDischarge.Plot
```

The following Figure \@ref(fig:ex-NY) for the New York daily discharge dataset displays a potential increasing trend over time.

```{r ex-NY, warning=FALSE, echo=FALSE, message=FALSE, fig.align='center', fig.cap='New York Discharge (cfs) 1990-2021'}
NY_DailyDischargePlot <- ggplot(NY_Discharge.processed, aes(x = Date, y = MeanDischarge)) +
  scale_y_log10() +
  geom_point(color = "turquoise4", size = 0.7) +
  geom_smooth(method = lm, color = "black") +
  ylab("Daily Discharge (cfs)") +
  labs(title = paste("Daily Discharge During Hurricane Seasons (1990-2021)"), subtitle = paste("Cold Spring Harbor, New York")) +
  scale_x_date(date_labels = "%Y", date_breaks = "2 year") +
  theme(axis.text.x = element_text(angle = 90)) 

NY_DailyDischargePlot
```


\newpage

# Analysis

The analysis of these datasets centers on creating time series objects of the discharge data. These time series objects are then decomposed to analyze the trend, seasonality, and remainder for the data values. New dataframes were then created with these three aspects of the time series. Within these dataframes, the trend and remainder values for each data point were combined to create a "non-seasonal" discharge value in order to better understand the actual changes in hurricane intensity and frequency during the hurricane season.

These non-seasonal values were then used to create another time series object which was graphed to visualize potential trends that may be occurring in the data as well as the spread of the discharge values. The spread of the discharge values indicates differences in discharge which is a proxy for hurricane intensity. To analyze trends statistically, rather than visually, a Mann Kendall test was performed. The Mann Kendall test produced a tau value which indicated whether the trend was increasing or decreasing. The tau values were compared across all three research locations in addition to p-values which displayed statistical significance of the results. 

### New York Analysis

The following plot displays the decomposed daily discharge components of the time series run on the New York dataset during hurricane season (Figure \@ref(fig:NY-time-series-analysis)).

```{r, NY-time-series-analysis, warning=FALSE, echo=FALSE, fig.align='center', fig.cap= "Decomposed Components of the New York Discharge Time Series"}

#NY Daily Discharge time-series
NY_DailyDischarge.ts <- ts(NY_Discharge.processed$MeanDischarge, start = c(1990,01), frequency = 153)
#head(NY_DailyDischarge.ts, 10) #idk if you need this kelsey!

#Decompose
NY_DailyDischarge.decompose <- stl(NY_DailyDischarge.ts, s.window = "periodic")
plot(NY_DailyDischarge.decompose)


#Nonseasonal MannKendall
NY_Components <- as.data.frame(NY_DailyDischarge.decompose$time.series[,2:3])
view(NY_Components)

#Add in a date column
NY_Components <- mutate(NY_Components,
                        Date = NY_Discharge.processed$Date)

#Adding non-seasonal component to dataframe
NY_Combine_Nonseasonal <- mutate(NY_Components, Nonseasonal = NY_Components$trend + NY_Components$remainder)
view(NY_Combine_Nonseasonal)

#Nonseasonal time series
NY_Nonseasonal.ts <- ts(NY_Combine_Nonseasonal$Nonseasonal, start= c(1990, 1), frequency = 153)

#MannKendall Nonseasonal Analysis
NY_Trend.Nonseasonal <- Kendall::MannKendall(NY_Nonseasonal.ts)
NY_Trend.Nonseasonal

#tau = 0.13; p-value = less than 0.05
#the null hypothesis of the test (i.e., Seasonal Mann Kendall) states that the data is stationary
#reject null hypothesis and we state that there is a trend

```
The Mann Kendall analysis performed on the New York non-seasonal time series produced a tau value of 0.13 with a p-value of less than 0.05. As a result, the data displays a significant, increasing trend for New York.


The plot below displays the non-seasonal daily discharge data in the New York time series which was produced by removing the seasonal component from the time series (Figure \@ref(fig:NY-time-series-analysis-continued)). 

```{r NY-time-series-analysis-continued, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap= "Nonseasonal Data from the New York Discharge Time Series"}

#Nonseasonal ggplot  

NY_NonSeasonal.Plot2 <- ggplot(NY_Combine_Nonseasonal, aes(x = Date, y = Nonseasonal)) +
  scale_y_log10() +
  geom_point(color = "turquoise4", size = 0.6) +
  labs(x = "Date", y = "Non-Seasonal Daily Discharge (cfs)", 
       title = "Non-Seasonal Mean Daily Discharge During Hurricane Seasons", 
       subtitle = "Cold Spring Harbor, New York") +
  geom_smooth(method = "lm", color = "black") +
  scale_x_date(date_labels = "%Y", date_breaks = "2 year") +
  theme(axis.text.x = element_text(angle = 90)) 
                               
print(NY_NonSeasonal.Plot2)

```



### North Carolina Analysis

```{r, NC-time-series-analysis, echo=FALSE, warning=FALSE, fig.align='center', fig.cap= "Decomposed Components of the North Carolina Discharge Time Series"}
# creating daily discharge time series
NewRiver.Daily.Discharge.ts <- ts(NewRiver.Discharge.PROC$MeanDailyDischarge_cfs,
                     start = c(1990, 06, 01), frequency = 153)

# decomposing daily time series
Daily.Discharge.decomp <- stl(NewRiver.Daily.Discharge.ts, s.window = "periodic")
plot(Daily.Discharge.decomp)

# adding components into a dataframe, removing seasonal component
Daily.Discharge.Components <- as.data.frame(Daily.Discharge.decomp$time.series[,2:3])

# adding trend and remainder into nonseasonal column
NonSeasonalDischarge.Daily <- mutate(Daily.Discharge.Components,
                                   Seasonality_Removed = Daily.Discharge.Components$trend +
                                     Daily.Discharge.Components$remainder)

# creating non-seasonal time series
NonSeasonal.DailyDischarge.ts <- ts(NonSeasonalDischarge.Daily$Seasonality_Removed,
                     start = c(1990, 06, 01), frequency = 153)

# adding date column to nonseasonal df
NonSeasonalDischarge.Daily <- NonSeasonalDischarge.Daily %>% 
                                mutate(Date = NewRiver.Discharge.PROC$Date)

# running Mann Kendall on time series with seasonality removed
MannKendall(NonSeasonal.DailyDischarge.ts)
```

Above is the decomposed time series of the North Carolina discharge data during hurricane months (Figure \@ref(fig:NC-time-series-analysis)). The Mann Kendall Test yielded a tau value of 0.0572 and a 2-sided p value of =< 2.22e-16.

```{r NC-time-series-analysis-cont, warning=FALSE, message=FALSE, fig.align='center', echo=FALSE, fig.cap= "Nonseasonal Data from the North Carolina Discharge Time Series"}
# creating nonseasonal plot
NonSeasonal.Plot <- ggplot(NonSeasonalDischarge.Daily, aes(Date, Seasonality_Removed)) +
                      geom_point(color = "forestgreen", size = 0.6) +
                      scale_y_log10() +
                      geom_smooth(method = "lm", color = "black") +
                      labs(x = "Date", y = "Non-Seasonal Daily Discharge (cfs)", 
                        title = "Non-Seasonal Mean Daily Discharge During Hurricane Seasons",
                        subtitle = "New River, North Carolina") +
                      scale_x_date(date_labels = "%Y", date_breaks = "2 year") +
                      theme(axis.text.x = element_text(angle = 90)) 
NonSeasonal.Plot
```

The plot above displays the non-seasonal data in the North Carolina time series which was produced by removing the seasonal component from the time series (Figure \@ref(fig:NC-time-series-analysis)). 

### Florida Analysis

Seen below is the decomposed time series of the Florida discharge data during hurricane months (Figure \@ref(fig:FL-anlaysis-time-series)). 

```{r FL-analysis-time-series, echo=FALSE, warning=FALSE, fig.align='center', fig.cap= "Decomposed plot of Florida discharge time series"}
#regular time series
florida_hurricane_ts <- ts(florida_hurricane_1990$Daily_Discharge_clean, start=c(1990,6), frequency = 153)  

#decomposed time series
florida_hurricane_Decomposed <- stl(florida_hurricane_ts, s.window = "periodic")
plot(florida_hurricane_Decomposed)
```


```{r FL-analysis-remove-seasonal, warning=FALSE, echo=FALSE}
#filtering out seasonal data
florida_hurricane_Decomposed_Components <- as.data.frame(florida_hurricane_Decomposed$time.series[,2:3])
florida_hurricane_Decomposed_Components <- florida_hurricane_Decomposed_Components %>%
  mutate(data = trend + remainder) 
#add date column
florida_hurricane_Decomposed_Components <- mutate(florida_hurricane_Decomposed_Components,
                        Date = florida_hurricane_1990$Date)

#run time series on new data with seasonal removed
nonseasonal_data_florida_ts <- ts(florida_hurricane_Decomposed_Components$data,
                                start=c(1990,6),
                                frequency=153) 
#look for trend and significance 
florida_trend <- MannKendall(nonseasonal_data_florida_ts)
florida_trend
```

From running a MannKendall nonseasonal trend analysis on the Florida discharge data set there is a significant overall decline in discharge over time with a tau value of -0.0516 (p = 7.7953e-08). Below this trend is visualized in a plot of the non-seasonal time series of Florida discharge with the significant decreasing trend shown with the black line (Figure \@ref(fig:FL-analysis-final-plot)). 

```{r FL-analysis-final-plot, warning=FALSE, message=FALSE, echo=FALSE, fig.align='center',fig.cap="Plot of nonseasonal time series of Florida discharge and overall trend"}
#plot of nonseasonal trend
FL_NonSeasonal.Plot <- ggplot(florida_hurricane_Decomposed_Components, aes(x = Date, y = data)) +
  scale_y_log10() +
  geom_point(color = "deeppink1", size = 0.6) +
  labs(x = "Date", y = "Non-Seasonal Daily Discharge (cfs)", 
       title = "Non-Seasonal Mean Daily Discharge During Hurricane Seasons", 
       subtitle = "North Prong St. Sebastian River, Florida") +
  geom_smooth(method = "lm", color = "black") +
  scale_x_date(date_labels = "%Y", date_breaks = "2 year") +
  theme(axis.text.x = element_text(angle = 90)) 
plot(FL_NonSeasonal.Plot)                               
```

### Comparison between Sites 

Each of the locations non-seasonal time series and trends are visualized and compared in the figure below (Figure \@ref(fig:combined-plot)). The color patterns are the same as above with Florida in pink, North Carolina in green, and New York in turquoise.  

```{r combined-plot, warning=FALSE, echo=FALSE, message=FALSE, fig.align='center', fig.cap='Non-Seasaonl Discharge Plot for all Gage Sites (cfs)'}
combined_NonSeasonal.Plot <- ggplot(NULL) +
  scale_y_log10(limits = c(0.25, 1200)) +
  geom_point(data = NonSeasonalDischarge.Daily, aes(Date, Seasonality_Removed), 
             color = "forestgreen", alpha = 0.05) +
  geom_smooth(data = NonSeasonalDischarge.Daily, aes(Date, Seasonality_Removed),
              method = "lm", color = "forestgreen") +
  geom_point(data = florida_hurricane_Decomposed_Components, 
             aes(x = Date, y = data), color = "deeppink1", alpha = 0.05) +
  geom_smooth(data = florida_hurricane_Decomposed_Components, 
             aes(x = Date, y = data), method = "lm", color = "deeppink1") +
  geom_point(data = NY_Combine_Nonseasonal, aes(x = Date, y = Nonseasonal), 
             color = "turquoise4", alpha = 0.05) + 
  geom_smooth(data = NY_Combine_Nonseasonal, aes(x = Date, y = Nonseasonal), 
              method = "lm", color = "turquoise4") +
  labs(x = "Date", y = "Non-Seasonal Daily Discharge (cfs)", 
       title = "Non-Seasonal Mean Daily Discharge During Hurricane Seasons") +
  scale_x_date(date_labels = "%Y", date_breaks = "2 year") +
  theme(axis.text.x = element_text(angle = 90))
plot(combined_NonSeasonal.Plot)
```

From this visual we can see that New York and North Carolina are seeing upward trends while Florida is seeing a downward trend in discharge. Also the overall discharge in New York is less than that of Florida and North Carolina. 

\newpage

# Summary and Conclusions

A poleward hurricane trend is apparent based on the daily discharge values associated with hurricane months across the East Coast of the U.S. Although discharge is noticeably greater in Florida and North Carolina, hurricane discharge trends are increasing in New York and North Carolina while the last 30 years yielded a slightly decreasing trend in Florida (Table \@ref(tab:table2)). The effect of poleward hurricane movement with increasing hurricane frequency/intensity could potentially lead to exacerbated natural disasters in northern states of the U.S. Large urban centers such as New York City are not equipped to handle such severe rain events. Given the substantial amount of impervious surface cover attributed to urbanization, future disastrous flooding events are inevitable (Coch 2015). Further research on the analysis could possibly include more sites along the East Coast to better comprehend the trends that may be occurring in terms of where hurricanes are becoming more/less frequent and intense. 


```{r table2, echo=FALSE}
State <- c("New York", "North Carolina", "Florida")
Tau <- c(0.13,0.0572,-0.0516)
P <- c("=< 2.22e-16","=< 2.22e-16","7.7953e-08")
table2 <- data.frame(State, Tau, P)
colnames(table2) <- c("State", "Tau value", "P-value")
  
knitr::kable(table2, caption ="List of tau values and p-values from non-seasonal time series analyses.")
```


\newpage

# References
Coch, N. K. 2015. Unique vulnerability of the New York–New Jersey metropolitan area to         hurricane destruction. Journal of Coastal Research 31:196–212.
Mudd, L., Y. Wang, C. Letchford, and D. Rosowsky. 2014. Assessing climate change impact on     the US East Coast hurricane hazard: temperature, frequency, and track. Natural Hazards     Review 15:4014001.
Schmeltz, M. T., S. K. González, L. Fuentes, A. Kwan, A. Ortega-Williams, and L. P. Cowan.     2013. Lessons from hurricane sandy: a community response in Brooklyn, New York. Journal     of urban health 90:799–809.

