---
title: "NC Discharge Analysis (New River)"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
```

Beginning by setting up the R session.

```{r include=FALSE}
# getting directory
getwd()

# loading packages
library(ggplot2)
library(tidyverse)
library(lubridate)
library(dplyr)
library(zoo)
library(Kendall)
library(ggfortify)
```

Next I load in the raw data and work to tidy the dataset including selecting columns, filtering for data post-1990, changing column names, filtering for hurricane season months, and parsing out the 'Date' column into its three components. This serves as the processed dataset. I then also created another dataframe for daily means of both 'Discharge_cfs' and 'GageHeight_ft'. Note: there are NA's for 'MeanGageHeight_ft' until 10/02/2007 as this is when this location was outfitted with this sensor.

```{r}
# loading in data
NewRiver.Discharge.RAW <- read.csv("./Data/Raw/NC_Discharge_NewRiver_RAW.csv")

# setting date column as date
NewRiver.Discharge.RAW$Date <- as.Date(NewRiver.Discharge.RAW$datetime, format = '%m/%d/%y')

# initial tidying of dataset
NewRiver.Discharge.RAW <- NewRiver.Discharge.RAW[-c(1), ] # removing non-data row

NewRiver.Discharge.PROC <- NewRiver.Discharge.RAW %>% 
                        select(agency_cd, site_no, 
                               X89345_00060, X89346_00065, Date) #selecting necessary columns

colnames(NewRiver.Discharge.PROC) <- c("Agency", "Site_No",
                                       "Discharge_cfs", "GageHeight_ft", "Date") # changing column names
  
# setting discharge and gage height columns as numeric
NewRiver.Discharge.PROC$Discharge_cfs <- as.numeric(NewRiver.Discharge.PROC$Discharge_cfs)
NewRiver.Discharge.PROC$GageHeight_ft <- as.numeric(NewRiver.Discharge.PROC$GageHeight_ft)

NewRiver.Discharge.PROC <- NewRiver.Discharge.PROC %>% 
                        group_by(Date) %>%  # grouping by date
                        mutate(Year = year(Date), # creating year column
                               Month = month(Date), # creating month column
                               Day = day(Date)) %>% # creating day column
                        filter(Year >= 1990) %>% # filtering for data 1990 or later
                        filter(Month > 5 & Month < 11) # filtering for data in hurricane season

# saving processd file as csv
write.csv(NewRiver.Discharge.PROC, "./Data/Processed/NC_Discharge_NewRiver_PROC.csv", row.names = FALSE)
                        
# creating daily discharge dataset
NewRiver.Daily.DischargeGage <- NewRiver.Discharge.PROC %>%  
            summarise(MeanDailyDischarge_cfs = mean(Discharge_cfs), # daily mean discharge
                      MeanDailyGageHeight_ft = mean(GageHeight_ft)) # daily mean gage height
```

Below is the time series analysis for the daily discharge 

```{r}
# initial data visualiation
DailyDischarge.Plot <- 
                ggplot(NewRiver.Daily.DischargeGage, aes(Date, MeanDailyDischarge_cfs)) +
                      geom_point(color = "darkgreen") +
                      geom_smooth(method = "lm", color = "black") +
                      labs(x = "Date", y = "Mean Daily Discharge (cfs)",
                        title = "Mean Daily Discharge During Hurrcane Seasons (1990-2021)", 
                           subtitle = "New River, North Carolina") +
                      theme(plot.title=element_text(hjust=0.5)) +
                      theme(plot.subtitle=element_text(hjust=0.5)) +
                      scale_y_log10()
DailyDischarge.Plot

# filling missing data with linear interpolation
NewRiver.Daily.Discharge.Clean <-
  NewRiver.Daily.DischargeGage %>%
  mutate(Discharge_cfs_Clean = zoo::na.approx(MeanDailyDischarge_cfs))

# creating daily discharge time series
NewRiver.Daily.Discharge.ts <- ts(NewRiver.Daily.Discharge.Clean$Discharge_cfs_Clean,
                     start = c(1990, 06, 01), frequency = 153)

# decomposing daily time series
Daily.Discharge.decomp <- stl(NewRiver.Daily.Discharge.ts, s.window = "periodic")
plot(Daily.Discharge.decomp)
```

Below is the analysis of the data with seasonality removed and the Mann Kendall test to determine trend change over time. 

```{r}
# adding components into a dataframe, removing seasonal component
Daily.Discharge.Components <- as.data.frame(Daily.Discharge.decomp$time.series[,2:3])

# adding trend and remainder for analysis
NonSeasonalDischarge.Daily <- mutate(Daily.Discharge.Components,
                                   Seasonality_Removed = Daily.Discharge.Components$trend +
                                     Daily.Discharge.Components$remainder)

# creating non-seasonal time series
NonSeasonal.DailyDischarge.ts <- ts(NonSeasonalDischarge.Daily$Seasonality_Removed,
                     start = c(1990, 06, 01), frequency = 153)

# adding date column to nonseasonal df
NonSeasonalDischarge.Daily <- NonSeasonalDischarge.Daily %>% 
                                mutate(Date = NewRiver.Daily.Discharge.Clean$Date)

# creating nonseasonal plot
NonSeasonal.Plot <- ggplot(NonSeasonalDischarge.Daily, aes(Date, Seasonality_Removed)) +
                      geom_point(color = "forestgreen", size = 0.6) +
                      scale_y_log10() +
                      geom_smooth(method = "lm", color = "black") +
                      labs(x = "Date", y = "Non-Seasonal Daily Discharge (cfs)", 
                        title = "Non-Seasonal Mean Daily Discharge During Hurricane Seasons",
                        subtitle = "New River, North Carolina") +
                      scale_x_date(date_labels = "%Y", date_breaks = "2 year") +
                      theme(axis.text.x = element_text(angle = 90)) 
NonSeasonal.Plot

# running Mann Kendall on time series with seasonality removed
MannKendall(NonSeasonal.DailyDischarge.ts)
```

