---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Insert title of project here"
subtitle: "Web address for GitHub repository"
author: "Elise Boos, Andrew Brantley, Kelsey Husted"
fontsize: 12pt
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(dygraphs)
library(xts)
library(ggfortify)

# Set your ggplot theme

my_theme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom", legend.text=element_text(size=8), 
        legend.title=element_text(size=10), panel.background = element_rect(fill = "gray95"), 
        panel.grid.major = element_line(size = 0.6, linetype = 'solid', colour = "white"))

theme_set(my_theme)

# Load your datasets
#New York Data
NY_Discharge <- read.csv("../Data/Raw/NY_MeanDailyDischarge.csv", stringsAsFactors = TRUE)

#North Carolina Data
NC_Discharge <- read.csv("../Data/Raw/NC_Discharge_NewRiver_RAW.csv", stringsAsFactors = TRUE)
```


# Rationale and Research Questions

Hurricanes are a serious natural disaster that often affects millions each year up and down the East Coast of the United States. Understanding how climate change is impacting the trends of hurricanes will be important information for strategic planning to protect both property and lives into the future. For this analysis we chose to look at discharge data of major rivers in states spanning the East Coast of the US in order to gain ome information on how locality and intensity of these hurricanes may be changing over the past three decades. 

We chose to investigate discharge data in Florida, North Carolina, and New York to gain an understanding of how hurricane regimes may be changing across the entire span of the coast. Furthermore, the comparative study will reveal if hurricanes are shifting poleward. Our central research questions revolved around investigating any increasing trends in discharge during the hurricane season (June-October) and where these increasing trends were more intense. 

## Research Questions

## Question 1: Has there been an overall increase in mean daily discharge during the Atlantic hurricane season from 1990 - 2021?

## Question 2: Has one portion of the East Coast of the US eperienced a disproportionate increase in hurricane activity, and its associated discharge, compared to other portions commonly affected by these hurricanes?

Are the general locations of hurricanes shifting towards the poles in the northern &/or southern hemisphere?

\newpage

# Dataset Information

All datasets used in this analysis were downloaded from the USGS website, specifically from their historical records of discharge data recorded by field dataloggers. The data was obtained from three states varying in latitude along the East Coast (i.e., Florida, North Carolina, and New York). Stream gage locations were selected based on proximity to major river basins and vulnerable counties susceptible to hurricane impacts. Additionally, Gage locations were chosen with sufficient data, going back to at least the 1990s. Similar timeline ranges for the datatsets ensures a time series analysis that is comparable among all three sites. Discharge data was the data type used in this analysis with all datasets having at minimum daily recordings but some having recordings as often as every 30 minutes. 


\newpage

# Exploratory Analysis 



\newpage

# Analysis

```{r}
NonSeasonal.Plot <- ggplot() +
  geom_point(aes(NY_Combine_Nonseasonal$Date, NY_Combine_Nonseasonal$Nonseasonal), 
             alpha = 1) + 
  geom_smooth(method = "lm", color = "turquoise4") +
  labs(x = "Date", y = "Non-Seasonal Daily Discharge (cfs)", 
       title = "Non-Seasonal Mean Daily Discharge During Hurricane Seasons", 
       subtitle = "All Research Locations") +
  scale_y_log10() +
  scale_x_date(date_labels = "%Y", date_breaks = "2 year") +
  theme(axis.text.x = element_text(angle = 90)) 
print(NonSeasonal.Plot)

```


\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
